ODGI: understanding pangenome graphs

### 2022-01-20 manuscript/main.tex
keywords

\usepackage[]{hyperref}
\hypersetup{colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black}
\usepackage{makecell}
\usepackage{comment}

@@

\abstract{
\textbf{Motivation:}
Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
These models offer the opportunity to study the entire genomic diversity of a population, including structurally complex regions.
Nevertheless, analyzing hundreds of gigabase-scale genomes using pangenome graphs is difficult as it is not well-supported by existing tools.
Hence, fast and versatile software is required to ask advanced questions to such data in an efficient way. \\
\textbf{Results:}
We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA \REVIEWED{pangenome graphs in the form of variation graphs}.
ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
Its fast parallel execution facilitates routine pangenomic tasks, as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
\textbf{Availability:}
ODGI is published as free software under the MIT open source license.
Source code can be downloaded from \url{https://github.com/pangenome/odgi} and documentation is available at \url{https://odgi.readthedocs.io}.
ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

\section{Introduction}
A pangenome models the full set of genomic elements in a given species or clade~\citep{cpang2018,Eizenga_2020}.
In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
%In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
\REVIEWED{A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
A bidirected sequence graph represents both strands of DNA.
On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{Garrison:2018}.
}
These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
Although these data structures are of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.

@@

\section*{Conflict of Interest}

The authors declare that they have no competing interests.

>>>>

\usepackage[draft]{hyperref}
\hypersetup{colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black}
\usepackage{makecell}
\usepackage{comment}

@@

\abstract{
\textbf{Motivation:}
Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
These models offer the opportunity to study the entire genomic diversity of a population, including structurally complex regions.
Nevertheless, analyzing hundreds of gigabase-scale genomes using pangenome graphs is difficult as it is not well-supported by existing tools.
Hence, fast and versatile software is required to ask advanced questions to such data in an efficient way. \\
\textbf{Results:}
We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA \REVIEWED{pangenome graphs in the form of variation graphs}.
ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
Its fast parallel execution facilitates routine pangenomic tasks, as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
\textbf{Availability:}
ODGI is published as free software under the MIT open source license.
Source code can be downloaded from \url{https://github.com/pangenome/odgi} and documentation is available at \url{https://odgi.readthedocs.io}.
ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
\\
\REVIEWED{\textbf{Keywords:} Navigating Variation Graphs. Pangenomic Liftover. Gigabase-Scale Pangenome Visualization. T2T Pangenome Graph Exploration.}
}

@@

\section{Introduction}
A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
%In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
\REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
A bidirected sequence graph represents both strands of DNA.
On this model, variation graphs add the concept of paths representing linear sequences as traversals through the nodes of the graph~\citep{Garrison:2018}. A path can be a genome, haplotype, contig, or read. [2,5]
}
These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
Although these data structures are of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.

@@

\REVIEWED{\section*{Conflict of Interest}}

\REVIEWED{The authors declare that they have no competing interests.}
